DP Problems of Interest

### 2012-10-23 docs/roadmap/benchmarks.tex
Pre-documentation-rework.

% --------------------------------------------------------------------------------
\subsubsection*{Intermediate results}
\begin{center}\begin{tabular}{cclrrr} \toprule
\bf Matrix & \bf Block & \bf Comment & \bf R(ms) & \bf T(ms) & \bf P(ms) \\ \midrule
1024 & 32 & CPU baseline & 1965.63 & 1191.13 & 6069.66 \\
1024 & 32 & GPU baseline & 838.67 & 499.87 & 516.82 \\
1024 & 32 & GPU sync improved & 642.37 & 316.40 & 343.29 \\

\bottomrule \end{tabular} \\[4pt]
Best timings for R=rectangle, T=triangle, P=parallelogram
\end{center}

>>>>

% --------------------------------------------------------------------------------
\subsubsection*{Intermediate results}
\begin{center}\begin{tabular}{cclrrr} \toprule
\bf Matrix & \bf Block & \bf Comment & \bf R(ms) & \bf T(ms) & \bf P(ms) \\ \midrule
1024 & 32 & CPU baseline & 1965.63 & 1191.13 & 6069.66 \\
1024 & 32 & GPU baseline & 838.67 & 499.87 & 516.82 \\
1024 & 32 & GPU sync improved & 642.37 & 316.40 & 343.29 \\
2048 & 32 & CPU baseline & 27229.28 & 15296.09 & 57323.04 \\
2048 & 32 & GPU P 32 blocks & 2864.27 & 1427.21 & 2096.34 \\
\bottomrule \end{tabular} \\[4pt]
Best timings for R=rectangle, T=triangle, P=parallelogram\footnote{$\le 32$ blocks on my GPU to prevent deadlock}
\end{center}

### 2012-10-23 docs/roadmap/problems.tex
Pre-documentation-rework.

% ----------------------------------------------------------------------------------------------------------------------------------------
\section{Introduction}
\subsection{Definitions}\ul
\item Block of computation: a block is simply a part of the DP matrix that we want to compute.
\item Wavefront: this is the place around which computation happen, typically. There should be some memory to store intermediate information between block of computations
\ule

\subsection{Problems classification}
In the literature, dynamic programming problems (DP) are classified according to two criteria:

@@

% ----------------------------------------------
\newpage
\subsection{Smith-Waterman (with gap extension at different cost)}\ol
\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

@@

		\item $\Sigma(BF)=\{stop, L, k\}$ with $0\le k < n$
		\ule
	\ule
\item Initialization:\ul
	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
	\ule
\item Recurrence:
\[\begin{array}{rcl}
W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	W_{(i+1,j)}+b & S\\
	W_{(i,j-1)}+b & W\\
	V_{(i,j)}+\delta(S_i,S_j) & V \\
	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
\end{array}\right\} = BW_{(i,j)}\\
\\
V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
	eh(i,j)+b \qquad\qquad\, \text{otherwise} & HL\\
	V_{(i+1,j-1)}+es(i,j) & IL \\
	VBI(i,j) & (i',j') \\
	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
\end{array}\right\} = BV_{(i,j)}\\
\\
F_{(j)}&=&\min\left\{\begin{array}{l|l}
	F_{(j-1)} & L \\ 
	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
\end{array}\right\} = BF_{(j)}\\
\\
VBI(i,j)&=&\min_{i<i'<j'<j}\{V_{(i',j')}+ebi(i,j,i',j')\} +c \quad|\quad (i',j')=BV_{(i,j)}
\end{array}\]

In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $W$,$V$, $VBI$ and $F$.

{\color{red}
\item Backtracking: Start at $BF_{(n)}$ using the recurrences
 \[\begin{array}{rcl}
	BF_{(j)} &=& \left\{\begin{array}{rcl} L&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
	\\
	BV(i,j) &=& \left\{\begin{array}{rcl}
		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
		IL &\implies& \big< {\rm stack}(i,j) \big> BV_{(i+1,j-1)} \\
		(i',j') &\implies& \big< \text{multi-loop from $(i,j)$ to }(i',j') \big> BV(i',j')\\
		k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
	\end{array}\right.\\
	\\
	BW_{(i,j)} &=& \left\{\begin{array}{rcl}
	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
	W & \implies & \big< bulge(j) \big> BW_{(i,j+1)} \\
	V &\implies& BV_{(i,j)} \\
	k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
	\end{array}\right.
\end{array}\]
}
\item Visualisation: \begin{center}\includegraphics[width=7cm]{problems_zucker.pdf}\end{center}
	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
\item Optimizations: {\color{red} XXX}
\ole

>>>>

% ----------------------------------------------------------------------------------------------------------------------------------------
\section{Problems classification}
\subsection{Definitions}\ul
\item Block of computation: a block is simply a part of the DP matrix that we want to compute.
\item Wavefront: this is the place around which computation happen, typically. There should be some memory to store intermediate information between block of computations
\ule

\subsection{Classification}
In the literature, dynamic programming problems (DP) are classified according to two criteria:

@@

% ----------------------------------------------
\newpage
\subsection{Smith-Waterman with affine gap extension cost}\ol
\item Problem: matching two strings $S$, $T$ with $|S_{\rm padded}|=m, |T_{\rm padded}|=n$.
\item Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

@@

		\item $\Sigma(BF)=\{stop, L, k\}$ with $0\le k < n$
		\ule
	\ule
\item Initialization:\ul
	\item Cost matrices: $W_{(i,i)}=V_{(i,i)}=0, F_{(0)}=$ energy of the unfolded RNA.
	\item Backtrack matrices: $BW_{(i,i)}=BV_{(i,i)}=BF_{(0)}=stop$.
	\ule
\item Recurrence:
\[\begin{array}{rcl}
W_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	W_{(i+1,j)}+b & S\\
	W_{(i,j-1)}+b & W\\
	V_{(i,j)}+\delta(S_i,S_j) & V \\
	\min_{i<k<j}W_{(i,k)}+W_{(k+1,j)} &k
\end{array}\right\} = BW_{(i,j)}\\
\\
V_{(i,j)}&=&\min\left\{\begin{array}{l|l}
	\infty \qquad\qquad\qquad\qquad {\rm if}(S_i,S_j) \text{ is not a base pair} & stop\\\\
	eh(i,j)+b \qquad\qquad\, \text{otherwise} & HL\\
	V_{(i+1,j-1)}+es(i,j) & IL \\
	VBI_{(i,j)} & (i',j') \\
	\min_{i<k<j-1}\{W_{(i+1,k)}+W_{(k+1,j-1)}\} +c & k
\end{array}\right\} = BV_{(i,j)}\\
\\
F_{(j)}&=&\min\left\{\begin{array}{l|l}
	F_{(j-1)} & L \\ 
	\min_{1\le i< j} (F_{(i-1)} + V_{(i,j)}) & i
\end{array}\right\} = BF_{(j)}\\
\\
VBI_{(i,j)}&=&\min\Big\{\min_{i<i'<j'<j}V_{(i',j')}+ebi(i,j,i',j')\} +c \,\,\Big|\,\, (i',j') \Big\}=BV_{(i,j)}
\end{array}\]

In practice, we don't go backward for larger values than 30, so we can replace $\min_{i<k<j}$ by $\min_{\max(i,j-30)<k<j}$ in the expressions of $VBI$ {\color{red} $W$,$V$ and $F$}.

\item Backtracking: Start at $BF_{(n)}$ using the recurrences
 \[\begin{array}{rcl}
	BF_{(j)} &=& \left\{\begin{array}{rcl} L&\implies& BF_{(j-1)} \\ i &\implies& BF_{(i-1)} + BV_{(i,j)} \end{array} \right.\\
	\\
	BV_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
		HL &\implies&\big< {\rm hairpin}(i,j) \big> \\
		IL &\implies& \big< {\rm stack}(i,j) \big> BV_{(i+1,j-1)} \\
		(i',j') &\implies& \big< \text{multi-loop from $(i,j)$ to }(i',j') \big> BV(i',j')\\
		k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
	\end{array}\right.\\
	\\
	BW_{(i,j)} &=& \color{red} \left\{\begin{array}{rcl}
	S & \implies & \big< bulge(i) \big> BW_{(i+1,j)} \\
	W & \implies & \big< bulge(j) \big> BW_{(i,j+1)} \\
	V &\implies& BV_{(i,j)} \\
	k &\implies& BW_{(i+1,k)} BW_{(k+1,j-1)}
	\end{array}\right.
\end{array}\]

\item Visualisation: \begin{center}\includegraphics[width=8cm]{problems_zucker.pdf}\end{center}
	% source: <<Parallization of dynamic programming recurrences in computational biology>> paper
\item Optimizations: {\color{red} XXX: we need to find how to encode $F_{(j)}$ such that we can normalize our computation into an existing recurrence shape.}
\ole
\newpage

% ------------------------------------------------------------------------------------------------
\section{Related problems}
